---
title: "Assignment-2: Key Trends in Azure Repositories and Topics"
author: "Divya Vemula"
date: "2024-11-03"
output: html_document
---

# Introduction

This report provides a comprehensive analysis of the **Azure** topic on GitHub, including repository characteristics, owner details, and trends in repository creation. 

**Topic Description**

- **Topic**: Azure 

- **Description**: A cloud computing platform and service created by Microsoft,
  offering various services including computing, analytics, storage, and
  networking. 

- **Link**: [Azure Topic on GitHub](https://github.com/topics/azure) 

- **Source**: GitHub 

- **Variables Included**: Personal access token for GitHub API

- **Total Repositories For Topic Azure is**: 16742 

- **Collected Repositories**: 100
  
# Data Wrangling

## Libraries Required

```{r setup, warning=FALSE, message=FALSE}
if (!require('tidyverse')) install.packages('tidyverse'); library('tidyverse')
if (!require('tidyr')) install.packages('tidyr'); library('tidyr')
if (!require('DALEX')) install.packages('DALEX'); library('DALEX')
if (!require('DALEXtra')) install.packages('DALEXtra'); library('DALEXtra')
if (!require('httr')) install.packages('httr'); library('httr')
if (!require('jsonlite')) install.packages('jsonlite'); library('jsonlite')
if (!require('stringr')) install.packages('stringr'); library('stringr')
if (!require('skimr')) install.packages('skimr'); library('skimr')
if (!require('recipes')) install.packages('recipes'); library('recipes')
if (!require('gh')) install.packages('gh');library(gh)
if (!require('kableExtra')) install.packages('kableExtra'); library('kableExtra')
if (!require('RColorBrewer')) install.packages('RColorBrewer'); library('RColorBrewer')
if (!require('ggplot2')) install.packages('ggplot2'); library('ggplot2')
if (!require('ggthemes')) install.packages('ggthemes'); library('ggthemes')

```

## Environment Variable Setup

**For security, we load the GitHub personal access token from an environment variable.**

```{r access token}
Sys.setenv(GITHUB_TOKEN = "")
token <- Sys.getenv("GITHUB_TOKEN")
```


**Checking the Repository Count for the Azure Topic**

```{r verifying Repository Count}
 topic <- "azure" 
 
 repo_count <- gh("GET /search/repositories", q = paste0("topic:", topic), 
                  per_page = 1, .token = token)$total_count 
 
 if (repo_count > 500) {
   cat(paste("The chosen topic", topic, "has more than 500 repositories.", "\n"))
   cat("Total Repositories For Topic Azure is:", repo_count, "\n")
 }else {
  cat(paste("The chosen topic", topic, "has fewer than 500 repositories.", "\n"))
 }
```

**Data Collection**

Fetching 100 repositories related to Azure

```{r 100 repositories}

repositories <- gh::gh("GET /search/repositories", 
                       q = "topic:azure", 
                       per_page = 100, .token = token)

```

**Data Cleaning and Transformation**

```{r Data Manipulation}
repos_df <- map_dfr(repositories$items, ~ data.frame(
  full_name = .x$full_name,
  created_at = .x$created_at %||% NA,
  size = .x$size %||% NA,
  language = .x$language %||% NA, 
  topics = ifelse(length(.x$topics) > 0, paste(.x$topics, collapse = ", "), "No Topics"), 
  stringsAsFactors = FALSE
))


repos_df <- repos_df |>
  drop_na() |>
  filter(!is.na(full_name) & !is.na(language) & 
         !is.na(topics) & !is.na(size)) |> 
  mutate(
    language = ifelse(language == "", NA, language),
    topics = ifelse(is.na(topics) | topics == "", "No Topics", topics)) |>
  distinct(full_name, .keep_all = TRUE) |>  
  arrange(size)


```

# Data Summary

## Table1: Repository Summary Table Sorted by Size

```{r Repository Summary}
repos_df  <-  head(repos_df,10)
caption_repo <- "<div style='background-color: #4CAF50; 
                    color: white; padding: 10px; text-align: center; 
                    font-size: 18px; font-weight:bold;'>RepositorySummary</div>"

repos_table <- kbl(repos_df, format = "html", caption =caption_repo ) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                  full_width = TRUE, position = "center") |>
    
    column_spec(1, width = "3em", background = "#e6f7ff") |> 
    column_spec(2, background = "#f0e68c") |>
    column_spec(3, background = "#e0f7fa") |>
    column_spec(4, background = "#e0f7fa", color = "darkgreen") |>
    column_spec(5, background = "#f0e68c", color = "darkgreen") |>
    row_spec(0, bold = TRUE, color = "white", background = "#4CAF50") |>
    row_spec(1:nrow(repos_df), color = "black")

  repos_table
```

**Table-1 Interpretation**

- **Diverse Languages**: The repositories use a variety of programming languages, including Go, Python, PowerShell, JavaScript, Java, C#, and TypeScript, reflecting a diverse technological stack in Azure-related projects.

- **Topics Coverage**: Some repositories lack specified topics, while others cover a broad range of subjects like authentication, cloud services, machine learning, and devops, indicating varied focus areas. 

- **Active Community**: The creation dates range from as recent as 2023 to older repositories started back in 2013, showcasing a long-standing and active community around Azure. 

- **Size Variation**: Repository sizes vary, with the smallest being 451 KB and the largest at 4167 KB, implying different levels of project complexity and scope. 

**These insights highlight the diversity and activity within the Azure-related repositories on GitHub.**


## Table2: Repository Owners Summary Table Sorted by Public Repository

```{r Repository Owners Summary}

get_user_data <- function(username, token) {
  user <- gh("GET /users/:username", username = username, .token = token)
  data.frame(
    login = user$login,
    public_repos = user$public_repos,
    created_at = user$created_at,
    followers = user$followers
  )
}

repos_data <- gh("GET /search/repositories", q = "topic:azure", per_page = 100, .token = token)
repos <- repos_data$items

owner_usernames <- unique(sapply(repos, function(x) x$owner$login))
owner_data <- bind_rows(lapply(owner_usernames, get_user_data, token = token))
owner_data <- arrange(owner_data, desc(public_repos))
caption_owners = "<div style='background-color: #A52A2A; 
                    color: white; padding: 10px; text-align: center; 
                    font-size: 18px; font-weight: bold;'>Repository Owners Summary</div>"
kbl(head(owner_data,10), format = "html", caption = caption_owners) |>
  
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = TRUE,
                position = "center",
                font_size = 15) |>
  row_spec(0, bold = TRUE, color = "white", background = "#A52A2A") |>
  row_spec(1:2, background = "#F5CBA7", color = "black") |>
  row_spec(3:4, background = "skyblue", color = "black") |>
  row_spec(5:6, background = "#F5CBA7", color = "black") |>
  row_spec(7:8, background = "skyblue", color = "black") |>
  row_spec(9:10, background = "#F5CBA7", color = "black")
```

**Table-2 Interpretation**

- **Diverse Contributors**: Both major organizations (Microsoft, Google) and individual contributors (Ramit Surana) play significant roles. 

- **High Influence**: Repositories from these owners indicate high activity and community influence, as reflected by the follower counts. 

- **Varied Focus Areas**: Contributions range from core cloud infrastructure to documentation and security, illustrating the breadth of the Azure ecosystem.


## Table3: Most Recently Closed Issues Table Sorted by Closed date


```{r Most Recently Closed Issues}
get_recent_closed_issue <- function(owner, repo, token) {
  issues <- tryCatch(
    gh("GET /repos/:owner/:repo/issues", owner = owner, repo = repo, state = "closed", .token = token),
    error = function(e) {
      message(paste("Error fetching issues for", owner, "/", repo, ":", e$message))
      return(NULL)
    }
  )
  if (is.null(issues) || length(issues) == 0) {
    return(data.frame(
      full_name = paste(owner, repo, sep = "/"),
      title = NA,
      created_at = NA,
      closed_at = NA,
      days_to_close = NA,
      stringsAsFactors = FALSE
    ))
  }
  
  recent_issue <- issues[[1]]
  title <- recent_issue$title %||% NA
  created_at <- recent_issue$created_at %||% NA
  closed_at <- recent_issue$closed_at %||% NA

  days_to_close <- if (!is.na(created_at) && !is.na(closed_at)) {
    as.numeric(difftime(ymd_hms(closed_at), ymd_hms(created_at), units = "days"))
  } else {
    NA
  }
  
  data.frame(
    full_name = paste(owner, repo, sep = "/"),
    title = title,
    created_at = created_at,
    closed_at = closed_at,
    days_to_close = days_to_close,
    stringsAsFactors = FALSE
  )
}

repos_data <- gh("GET /search/repositories", q = "topic:azure", per_page = 100, .token = token)
repos <- repos_data$items
repo_list <- lapply(repos, function(x) list(owner = x$owner$login, repo = x$name))
issue_data <- bind_rows(lapply(repo_list, function(x) get_recent_closed_issue(x$owner, x$repo, token)))
issue_data <- issue_data |> arrange(desc(closed_at))
caption_issues = "<div style='background-color: #4CAF50; 
                    color: white; padding: 10px; text-align: center; 
                    font-size: 18px; font-weight: bold;'>Most Recently Closed Issues for Azure Repositories</div>"

kable(head(issue_data, 10), format = "html", caption = caption_issues) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) |>
  column_spec(1, bold = TRUE, color = "darkblue", background = "#c8e6c9") |>   
  column_spec(2, background = "#a5d6a7") |>                                
  column_spec(3, background = "#81c784") |>
  column_spec(4, bold = TRUE, color = "darkblue", background = "#c8e6c9",
              border_left = TRUE, width = "8em") |> 
  column_spec(5, background = "#81c784") |>   # Action button style
  row_spec(0, bold = TRUE, color = "white", background = "#DC143C")   

```

**Table-3 Interpretation**

- **Efficiency:** Most issues were resolved quickly, reflecting efficient issue handling.

- **Diverse Focus Areas:** Issues addressed range from bug fixes and enhancements to support queries and test configurations.

- **Azure SDKs:** Multiple issues closed are related to Azure SDKs, indicating ongoing improvements and support.

**These insights highlight the responsiveness and diverse focus areas in managing Azure repository issues.**


# Data Visualizations

## Plot1 Bar Plot for Top 5 Co-occurring Topics

```{r plotting Bar Plot, fig.align="left", fig.width = 14, fig.height=8}

topics_list <- c()
for (repo in repositories$items) {
  if (!is.null(repo$topics) && length(repo$topics) > 0) {
    topics_list <- c(topics_list, paste(repo$topics, collapse = ", "))
  }
}

topics_df <- data.frame(
  topic = unlist(strsplit(paste(na.omit(topics_list), collapse = ", "), ", ")), 
  stringsAsFactors = FALSE
)

top_5_topics <- topics_df |>
  filter(topic != "azure") |>
  count(topic, sort = TRUE) |>
  head(5)

plt1 <- ggplot(top_5_topics, aes(x = reorder(topic, n), y = n, fill = topic)) + 
  geom_bar(stat = "identity", color = "black", linewidth = 0.7) +
  scale_fill_brewer(palette = "Spectral") +
  labs(title = "Top 5 Topics Co-occurring with Azure",
       x = "Topic",
       y = "Count",
       fill = "Topic") +
  theme_minimal(base_size = 15) + 
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
        axis.text.x = element_text(size = 12, face = "italic"),
        panel.background = element_rect(fill = "lightyellow", color = NA)) +
  coord_flip()
suppressWarnings(print(plt1))

```

**Pattern Inference:**

**Key Insights from "Top 10 Topics Co-occurring with Azure"**

  - **AWS**: 35 occurrences, strong interconnections in cloud discussions.
  
  - **GCP**: 25 occurrences, highlights competitive and complementary    
  relationships.
  
  - **Cloud**: 23 occurrences, emphasizes Azure's role in the broader
  ecosystem.
  
  - **Kubernetes**: 20 occurrences, suggests significant use in
  containerization.
  
  - **Hacktoberfest**: 18 occurrences, reflects support for community-driven
  events.

**Interpretation**

  - **Cloud Ecosystem**: Azure's integral role highlighted by frequent mentions
  of AWS, GCP, and Cloud.

  - **Containerization**: Kubernetes mentions indicate Azure's relevance in
  modern development practices.
  
  - **Community Engagement**: Hacktoberfest shows Azure's participation in the
  open-source community.


## Plot2: Line Chart for Repository Creation Over Time

```{r plotting Line Chart Plot, fig.align="left", fig.width = 14, fig.height=8}
repos_time <- map_dfr(repositories$items, ~ data.frame(
  full_name = .x$full_name,
  created_at = .x$created_at %||% NA,
  size = .x$size %||% NA,
  language = .x$language %||% NA, 
  topics = ifelse(length(.x$topics) > 0, paste(.x$topics, collapse = ", "), "No Topics"), 
  stringsAsFactors = FALSE
))

repos_time$created_at <- as.Date(repos_time$created_at)
repos_time$year <- format(repos_time$created_at, "%Y")
repos_yearly <- repos_time |>
  group_by(year) |>
  summarize(count = n()) |>
  arrange(year)

repos_yearly$legend_label <- paste(repos_yearly$year, "(", repos_yearly$count, ")", sep = "")
plt2 <- ggplot(repos_yearly, aes(x = as.integer(year), y = count, group = 1)) +
  geom_line(aes(color = factor(legend_label)), linewidth = 2) +  
  geom_point(aes(color = factor(legend_label)), size = 3) + 
  scale_color_viridis_d(option = "C", direction = -1) + 
  labs(
    title = "Trend of Repository Creation Over Time",
    x = "Year",
    y = "Number of Repositories",
    color = "Year"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title.x = element_text(face = "bold", size = 15),
    axis.title.y = element_text(face = "bold", size = 15),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    panel.background = element_rect(fill = "#d1e7dd", color = NA),
    plot.background = element_rect(fill = "#eef5db", color = NA), 
    panel.grid.major = element_line(linewidth = 0.1, linetype = 'solid', colour = "gray"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(
    min(as.integer(repos_yearly$year)),
    max(as.integer(repos_yearly$year)),
    by = 1
  ))
suppressWarnings(print(plt2))
```


**Patern Inference:**

**Key Insights**

  - **Initial Low Activity (2010-2013)**: Minimal repository creation. 

  - **Significant Growth (2014-2017)**: Rapid increase, peaking in 2017. 

  - **Fluctuations Post-2017**: Varied creation rates, with a notable rise in     2023. 

**Inference**

  - **Growth and Interest**: Significant growth period indicates heightened         interest in Azure. 

  - **Market Dynamics**: Fluctuations reflect market changes and technology         trends. 

  - **Recent Trends**: Renewed interest or stabilization from 2021 to 2023.

# Conclusion

- This report provides a comprehensive analysis of the Azure topic on GitHub, including repository characteristics, owner details, and trends in repository creation.
- Azure’s popularity is evident from the consistent growth in repository creation over time and its common co-occurrence with topics like AWS and GCP.









